Exploring the ages of rotating stars using galactic dynamics: a novel approach to calibrating gyrochronology

### 2019-10-14 paper/results.tex
some improvements but many more needed

To explore the relationship between rotation period, \teff\ and age, we
selected groups of stars within different gyrochronal age ranges \citep[where
age was calculated using the][gyrochronology relation]{angus2019}, and
calculated the velocity dispersion (\sigmavb), the standard deviation of \vb\
velocities, as a function of effective temperature for each age group.
We performed 3$\sigma$ sigma-clipping on the stellar velocities in each age
and temperature bin to remove non-Gaussian outliers.
Without sigma-clipping, we found that a small number of high velocity
outliers at the low-temperature end of our sample substantially raised the
velocity dispersion for cooler stars.
Ages were calculated using dereddened \gaia\ \gcolor\ color, however
throughout this paper we show rotation periods as a function of effective
temperature, \teff.
We chose to use \teff\ as the abscissa because it is easier to divide stars
into bins of roughly equal numbers in \teff-space than in color-space.

@@

(excluding visual binaries and subgiants) in grey, with coeval groups shown in
color.

@@

dispersion and rotation periods increase with age as expected.

@@

If the gyrochronology relations worked at all ages and temperatures, the
bottom panel of figure \ref{fig:age_cut} would show a flat relationship
between velocity dispersion and \teff at all ages.

@@

The lines of constant age (isochrones) sweeping diagonally upwards in the top
panel of figure \ref{fig:age_cut} are too steeply sloped at old ages.

@@

\begin{figure}
  \caption{
      Top: Rotation period vs effective temperature for stars in the \mct\
    sample, colored by the velocity dispersions of stars calculated over a
    grid in $\log_{10}$(period) and \teff.
    Bottom: the velocity dispersions of groups of stars, shown as a solid
    grid for clarity.
    The hatched area indicates the temperature regime where selection biases
    could play an important role, so these velocity dispersions should be
    interpreted with caution.
    The black solid lines on both panels show a 1.1 Gyr isochrone, calculated
    with the \citet{angus2019} gyrochronology relation, which roughly traces
    the rotation period gap.
    The black dashed lines show a 650 Myr isochrone, indicating the location
    and shape of the Praesepe cluster (to which this gyrochronology model was
    calibrated).
    The black points show the 1.1 Gyr NGC 6811 open cluster.
}
  \centering
    \includegraphics[width=1\textwidth]{vplot}
\label{fig:dispersion_period_teff}
\end{figure}
Figure \ref{fig:dispersion_period_teff} shows rotation period vs. \teff\ for
our sample, coloured by (\vb) velocity dispersion, where \sigmavb\ was
calculated for groups of stars over a grid in $\log_{10}$(period) and
temperature.
If we assume that mass dependent heating does not strongly affect this sample
and \vb\ at low galactic latitudes is an unbiased tracer of \vz, then \vb\
velocity dispersion can be interpreted as an age proxy and stars plotted in a
similar color in figure \ref{fig:dispersion_period_teff} are similar ages.
Interpreted this way, lines of constant age (isochrones) appear to follow the
shape of the Praesepe-based gyrochronology model (black solid and dotted
lines) at young ages.
However, at older ages it appears that the relation between rotation period
and \teff\ flattens out, until eventually rotation period {\it decreases} with
decreasing effective temperature at a given age.

The shape of the period-\teff\ relations at old ages appears to follow the
shape of the upper detection edge.
The `M dwarf dip' \citep{vansaders2018}, a feature of the \mct\ rotation
period catalog characterized by a dearth of slowly rotating M dwarfs between
$\sim$2750 and 4250 K, is reflected in the lines of constant velocity
dispersion (and presumed age) in the top panel of figure
\ref{fig:dispersion_period_teff}.
If the shape of the upper edge of rotation period measurements is created by a
detection limit, it could indicate the rotation periods at which stars of
different temperatures become relatively inactive (and their rotation periods
therefore become undetectable).
% Once inactive, stars would have little photometric variability induced by
% star spots and their rotation periods would be difficult to measure, so may
% not feature in the \mct\ catalog.
If so, figure \ref{fig:dispersion_period_teff} suggests that stars cooler than
$\sim$4500 K become inactive at a similar age, and stay active longer than
stars hotter than $\sim$4500 on average.

@@

>>>>

To explore the relationship between rotation period, \teff\ and age for field
K dwarfs, we calculated gyrochronal ages using dereddened \Gaia\ photometry
($G$, $G_{BP}$ and $G_{RP}$) and rotation periods from \citet{mcquillan2014}.
We used the \citet{angus2019} gyrochronology relation which is a simple,
separable relation in \gcolor\ color and age, calibrated using the
period-color relation of Praesepe and the period-age relation of Praesepe and
the Sun.
The large number of Praesepe members with precise rotation periods from the
\ktwo\ mission \citep{howell2014, douglas2017, rebull2017}, spanning spectral
types F through early M, makes it a good cluster for calibrating the
period-color relation of stars at young ages.
However, this is an extremely simple model and although it accurately
describes the rotation periods of F and G stars up to around 2.5 Gyr (the age
of NGC 6819 -- the oldest cluster with available rotation periods), it
over-predicts the rotation periods of K dwarfs in the 1.1 Gyr NGC 6811
cluster.
These gyrochronal ages are shown on a \gaia\ color-magnitude diagram (CMD) in
figure \ref{fig:age_gradient}.
The stars with old gyrochronal ages, plotted in yellow hues, predominantly lie
along the upper edge of the MS, where stellar evolution models predict old
stars to be, however the majority of these `old' stars are bluer than \gcolor\
$\sim$ 1.5 dex.
This suggests that either old M dwarfs are missing from the \mct\ catalog, or
the \citet{angus2019} gyrochronology relation under-predicts the ages of
low-mass stars.
Given that lower-mass stars stay active for longer than higher-mass stars
\citep[\eg][]{west2011}, and are therefore more likely to have measurable
rotation periods at old ages, the latter scenario seems more likely.

To further explore the relationship between rotation period, \teff\ and age,
we first removed high and low velocity outliers by performing 3$\sigma$
sigma-clipping on the \vb\ velocities.
Without sigma-clipping, we found that a small number of high velocity
outliers at the low-temperature end of our sample substantially raised the
velocity dispersion for cooler stars.
We then selected groups of stars within different gyrochronal age ranges and
calculated the standard deviation of \vb\ velocities (\sigmavb), as a function
of effective temperature for each age group.
Ages were calculated using dereddened \gaia\ \gcolor\ color, however
throughout this paper we show rotation periods as a function of effective
temperature, \teff, because it is easier to divide stars into bins of roughly
equal numbers in \teff-space than in color-space.

@@

(excluding visual binaries and subgiants) in grey, with coeval groups
\citep[according to the][relation]{angus2019} shown in color.

@@

dispersion and rotation period increases with age as expected.

@@

If \citet{angus2019} gyrochronology relation worked at all ages and
temperatures, the bottom panel of figure \ref{fig:age_cut} would show a flat
relationship between velocity dispersion and \teff at all ages.

@@

% The lines of constant age (isochrones) sweeping diagonally upwards in the top
% panel of figure \ref{fig:age_cut} are too steeply sloped at old ages.

@@

\begin{figure}
  \caption{
      Top: Rotation period vs effective temperature for stars in the \mct\
    sample, colored by the velocity dispersions of stars calculated over a
    grid in $\log_{10}$(period) and \teff.
    Bottom: the velocity dispersions of groups of stars, shown as a solid
    grid for clarity.
    The hatched area indicates the temperature regime where selection biases
    could play an important role, so these velocity dispersions should be
    interpreted with caution.
    The black solid lines on both panels show a 1.1 Gyr isochrone, calculated
    with the \citet{angus2019} gyrochronology relation, which roughly traces
    the rotation period gap.
    The black dashed lines show a 650 Myr isochrone, indicating the location
    and shape of the Praesepe cluster (to which this gyrochronology model was
    calibrated).
    The black points show the 1.1 Gyr NGC 6811 open cluster.
}
  \centering
    \includegraphics[width=1\textwidth]{vplot}
\label{fig:vplot}
\end{figure}
Figure \ref{fig:dispersion_period_teff} shows rotation period vs. \teff\ for
our sample, coloured by (\vb) velocity dispersion, where \sigmavb\ was
calculated for groups of stars over a grid in $\log_{10}$(period) and
temperature.
If we assume that mass dependent heating does not strongly affect this sample
and \vb\ at low galactic latitudes is an unbiased tracer of \vz, then \vb\
velocity dispersion can be interpreted as an age proxy and stars plotted in a
similar color in figure \ref{fig:dispersion_period_teff} are similar ages.
Interpreted this way, lines of constant age (isochrones) appear to follow the
shape of the Praesepe-based gyrochronology model (black solid and dotted
lines) at young ages.
However, at older ages it appears that the relation between rotation period
and \teff\ flattens out, until eventually rotation period {\it decreases} with
decreasing effective temperature at a given age.

The shape of the period-\teff\ relations at old ages appears to follow the
shape of the upper detection edge.
The `M dwarf dip' \citep{vansaders2018}, a feature of the \mct\ rotation
period catalog characterized by a dearth of slowly rotating M dwarfs between
$\sim$3750 and 4250 K, is reflected in the lines of constant velocity
dispersion (and presumed age) in the top panel of figure
\ref{fig:dispersion_period_teff}.
If the shape of the upper edge of rotation period measurements is created by a
detection limit, it could indicate the rotation periods at which stars of
different temperatures become relatively inactive (and their rotation periods
therefore become undetectable).
% Once inactive, stars would have little photometric variability induced by
% star spots and their rotation periods would be difficult to measure, so may
% not feature in the \mct\ catalog.
If so, figure \ref{fig:dispersion_period_teff} suggests that stars cooler than
$\sim$4500 K stay active longer than stars hotter than $\sim$4500 on average.

@@

\begin{figure}
  \caption{
    Similar to figure \ref{fig:vplot}, with mass instead of \teff\ on the
    x-axis.
    Masses are from the \kepler\ input catalog.
    The white lines show the \citet{spada2019} rotational evolution models at
    0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr, where age increases with rotation
    period.
    These models include age and mass-dependent coupling between the stellar
    core and envelope.
    The oldest models (4 and 4.57 Gyrs), at the largest rotation periods, show
    an inversion at $\sim$0.7-0.5 M$_\odot$, where rotation period briefly
    decreases with decreasing mass.
    A similar phenomenon is visible in the velocity dispersions of field stars
    shown as colored points in the background of this figure.
}
  \centering
    \includegraphics[width=1\textwidth]{masses_with_models}
\label{fig:masses_with_models}
\end{figure}
